Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-16 manuscript/manuscriptRW.Rmd
no idea what has changed here but I save it just in case
small additions manuscript

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

In spite of more or less substantial mismatches between model results
and data, it is remarkable that most of the models capture the essential
features of the highly complex process of tree establishment, while not
having been parameterized with such data. We conclude that in spite of
all the uncertainties, a lot can be gained by a focus on the modeling of
establishment processes in these model, this will make their long-term
forest dynamics projections under changing environmental conditions much
more robust.

@@

More specifically, we aimed to answer the following questions: (1) Are
models of forest dynamics capturing accurately tree ingrowth levels,
initial tree diversity and mortality in tree establishment? (2)Do model
traits explain differences in model performance? (3) How well do models
capture total ingrowth and individual species regeneration niches across
environmental gradients of light availability, temperature, and soil
moisture?

@@

The starting point for the models establishment differs as well, ranging
from seedbank, seed or seedling to sapling (i.e., trees much larger than
10 cm height, often at least 2-3 m tall). Models that start from seed
need to include a larger number of ecological processes such as
germination and survival of young seedlings, whereas models that started
from saplings have to aggregate via parameterizations several ecological
processes because they are not treated explicitly. This reduces model
complexity but comes at the cost of blurring process representation. In
the models used in this study the overall complexity in the
establishment modules varies considerably. Following @bugmann2022, we
can classify seven models (iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS,
ForClim and TreeMig) as having rather high complexity in their
establishment modules (mean establishment formulation complexity
complexity across all processes \>0, Table \@ref(tab:tableModels)).

Two models, ForClim and ForCEEPS, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a different approach where
individual species properties and its relation to the environment are
more important [@huber2020]. These two model variants allow us to
evaluate a more process-based and complex module (variant 11) against a
simple module (variant 1), while the rest of the model structure is
identical. Similarly, the two ForCEEPS variants allow us to isolate the
importance of the canopy feedback (i.e. simulated actual composition and
relative abundance of species in the plot) via seed trees for the
quantity and quality of simulated ingrowth, as one variant includes this
feedback whereas the other does not.

@@

kable(modelDF, 
      format = "latex", booktabs = TRUE,
      caption = "List of models included.") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |> 
      kableExtra::column_spec(column = 2, width = "3em") |> 
      kableExtra::column_spec(column = 3, width = "3em") |> 
      kableExtra::column_spec(column = 4, width = "4em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "5em") |>
      kableExtra::column_spec(column = 7, width = "5em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "8em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "5em") |>
      kableExtra::column_spec(column = 12, width = "8em") |>
      kableExtra::add_footnote(c("S-Stand, L-Landscape, G-Global",
                                 "PB-Process based, EM-Empirical", 
                                 "C-Cohort, I-Individual",
                                 "RG-Regeneration, RC-Recruitment",
                                 "Y-Yes, N-No",
                                 "D-Daily, M-Monthly"),

@@

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, water
balance, soil quality and exposition (i.e., slope and aspect). Each of
these sites featured at least two consecutive measurements. Ingrowth
thresholds for the sites differed between 0 and 10 cm, we defined two
data sets where one included all sites with a dbh threshold of 7cm or
lower and the other included another 35 sites with dbh thresholds
between 7 and 10 cm. Thus we focused on two data sets with the
harmonized dbh thresholds of 7 cm and 10 cm.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species diversity of
the ingrowth, and ingrowth mortality. We evaluated the species diversity
across models and in relation to the observed data by calculating the
Shannon index based on the relative proportion of the species in terms
of basal area. The Shannon index was calculated for the establishment
($H_{n,r}$) (Equation \ref{eqpiShannonR}) and at the stand level
($H_{n,s}$)(Equation \ref{eqpiShannon}). The higher the value of the
index, the higher the species diversity at a particular site and sample.

```{=tex}
\begin{equation}
H_{n,r} = - \sum_{i = 1}^{s}{p_{n,i}\ln(p_{n,r,i})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,r,i} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{n,s} = - \sum_{i = 1}^{s}{p_{n,i}\ln(p_{n,i})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{n,s,i} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and have basal area
above zero in sample $n$; $p_{n,r,i}$ is the proportion of the species
$i$ calculated as the ingrowth basal area ($r_{BA_{n,i}}$) for that
species $i$ relative to the recruited basal area ($TotalR_{BA_{n}}$) of
the sample $n$; $p_{n,s,i}$ is the proportion of the species $i$
calculated as the basal area ($BA_{n,i}$) for that species $i$ relative
to the total basal area ($TotalR_{BA_{n}}$) of the sample $n$.

Mortality in tree establishment was assessed based on the ratio of
ingrowth between the 7 and 10 cm diameter thresholds. We used the
Reineke self-thinning rule [@reineke1933; @pretzsch2003] as a reference
to estimate whether the different ingrowth values were above or below
the expected theoretical rate between these two diameters. The Reineke
self-thinning rule is usually calculated from even-aged, single species
stands and assumes a relationship between the number of stems and the
quadratic mean diameter in fully stocked pure stands. The value used in
our comparisons was 1.77 (i.e., we expect stem numbers at 7cm to be 77%
higher than at 10cm), calculated using Equation \ref{eqReineke}.

```{=tex}
\begin{equation}
Reineke = (7/10)^{(-1.605)} \label{eqReineke}
\end{equation}
```
We assessed model traits in relation to model performance focusing on
features such as model complexity as defined in @bugmann2022, model type
(empirical or process based), models with and without canopy feedback
(the level and species composition of the ingrowth is influenced by the
existing adult tree community) and the models scale (stand, landscape or
global). We tested for significant differences considering each of these
model features and between observed and simulated ingrowth
overestimation, ingrowth levels and species diversity respectively.

The total ingrowth and individual species regeneration niches were
evaluated across environmental gradients of light availability,
temperature, and soil moisture. Ingrowth levels were evaluated along
three key environmental gradients, i.e., stand basal area (the basal
area of all the trees in each sample including tree establishment) as a
proxy of light availability at the forest floor, the annual degree-day
sum (DD) [@allen1976; @fischlin1995] as a proxy for growing season
warmth, and the climatic water balance (CWB) as a proxy for soil
moisture [@speich2019]. The mean ingrowth values were calculated as the
mean across the 200 samples per site. The observed data were modeled
using a Generalized Additive Model [@wood2011] with a negative binomial
distribution and restricted maximum likelihood to better understand the
environmental gradients trend in the levels of the observed ingrowth,
relative to the simulation results per model.

@@

```{r over, echo=FALSE, fig.cap="Mean ingrowth levels across all samples per site, plotted for the 200 sites and for each model. The red dashed lines shows the 25th and 75th percentiles for the 7cm diameter threshold in the observed data. There are two boxplots for each model where the left lighted boxplot corresponds to 7cm and the right darker boxplot to 10cm", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

@@

```{r H710, echo=FALSE, fig.cap="Shannon index of tree ingrowth, calculated by basal area, as the mean value across the 200 samples per site for the diameter thresholds of 7 and 10 cm, respectively. The red dashed lines represent the 25th percentile of the observed data for the threshold of 7 cm and the 75th percentile of the observed data for the threshold of 10 cm.", out.width='100%'}
    knitr::include_graphics(here::here("figures", "H7_10.png"))
```

@@

For both the observed and the simulated data, and both ingrowth
thresholds, species diversity in the ingrowth compared to stand-level
diversity did not have clear patterns across all the models (Figures S1
and S2). Most models captured reasonably well or overpredicted species
diversity at the stand level (Figure \@ref(fig:combinedRichness)A,
center; cf. Figure S3), but a characteristic feature was that
simulations had a much lower variability of diversity compared to
observations. Models overpredicting species diversity included ForClim1,
ForClim 11, FORMIND and TreeMig (Figure \@ref(fig:combinedRichness)A,
left). Several models underpredicted species diversity, i.e., LPJ-GUESS,
ForCEEPS(f) and 4C (Figure \@ref(fig:combinedRichness)A, right).

The majority of the models overpredicted ingrowth diversity and only two
of them underestimated ingrowth diversity (4C and Landis-II) (Figure
\@ref(fig:combinedRichness)A and C, cf. Figure S4). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) overestimate, underestimated and captured
reasonably well ingrowth species diversity compared to the observations.

@@

There were strong differences among the models regarding the mortality
rate between 7 and 10 cm (Figure \@ref(fig:mortality11)). Most of the
models featured a mortality rate significantly larger than the observed
data, thus at least partially compensating for the general
overestimation of ingrowth levels (Figure \@ref(fig:over) and Table S5);
six models did not feature a significant difference compared to the
observed data (ForCEEPS, FORMIND, ForClim 1 and 11, iLand and aDGVM2).
Four models (xComp, PICUS, LandClim and TreeMig) featured very high
mortality rates (i.e., well above the Reineke self-thinning line), which
compensated for the strong initial overestimation of ingrowth (cf.
Figure \@ref(fig:over)). Interestingly, two models that underestimated
overall ingrowth levels (SIBYLA and Landis-II; cf. Figure
\@ref(fig:over)) featured mortality rates that were close to but still
above the self-thinning line (Figure \@ref(fig:mortality11)).

@@

There was no significant relationship between the mean complexity of the
establishment module (Table \@ref(tab:tableModels)) and the
overestimation proportion at a diameter threshold of 7 cm (Figure S7).
All the models except 4C had significant differences between the
observed mean ingrowth values and the simulated, therefore it was not
possible to differentiate in terms of performance between models forest
type (empirical or process based) and across models scales (stand,
landscape or global) (Table S5). The ingrowth species diversity values
simulated per models were also assessed againts the observed data to see
if there was a difference between models that include feedback compare
to those that did not include feedback, all the models except
ForCEEPs(f) and FORMIND had significant differences (Table S6). It is
noteworthy the change from overpredicted species diversity in the
ingrowth for the model ForCEEPS to a closer diversity levels when the
recruitment module included feedback (ForCEEPs(f)). Again, simulated
diversity was considerably less variable than observed diversity.

@@

The observed ingrowth did not change much across the climatic water
balance gradient, and showed a slight increase with temperature. The
models did not match the observed ingrowth patterns across gradients of
soil moisture and temperature (Figure
\@ref(fig:)\@ref(fig:nvTrend710BAlightTemp) and S7), and neither was
there a clear pattern across models. Regarding the soil moisture
gradient, a group of models featured decreasing ingrowth with increasing
soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp, PICUS, LandClim,
TreeMig and iLand), while a few models showed the opposite trend
(ForCEEPS(f), FORMIND and LPJ-GUESS). Across the temperature gradient,
all the models that featured a decrease in ingrowth levels with
increasing soil moisture showed the opposite trend (i.e., more ingrowth
with increasing temperature), except for 4C. The other models did not
feature a clear pattern.

@@

Correctly capturing species composition in tree establishment is
important to assess the future functional diversity of a forest, e.g.,
its sensitivity to drought or resilience to disturbances [@redmond2015;
@seidl2022]. In the simulations, overall ingrowth species diversity
levels were well within the observed range for two thirds of the models
(10 out of 15). Thus, while most of the models are facing difficulties
to quantitatively match ingrowth levels (cf. above), their performance
is much better when it comes to matching the composition of ingrowth as
a function of abiotic and biotic conditions. Most models maintained or
even increased ingrowth diversity between the 7 and 10 cm thresholds,
and the same is visible from the empirical data, but the differences are
not significant. Diversity variations may be explained by the way the
individual models consider establishment processes [@konig_tree_2022],
e.g., in the absence of adults, ingrowth can be linked to either seed
rain-dispersal, or to a seed/seedling bank.

The species diversity of the entire stand was better captured by the
models than the diversity of tree ingrowth, and only a few models
overpredicted stand-level diversity. Defining ingrowth composition as
being proportional to adult abundance, regardless of productivity, might
be a simple and conservative assumption to safely maintain relative
species abundances [@hanbury-brown2022], although this may be
simplistic. Furthermore, based on the simulation results, there is no
evidence that models with feedback captured better the ingrowth species
diversity than those without feedback. The similar performance of models
with and without feedback is likely because the models put more weight
to the regeneration niche arising from abiotic and biotic filters, than
from the habitat niche of the adult trees [@grubb1977].

@@

Funding of the workshop was provided by the COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Grant ETH-35 18-1). ODY was funded by SNF project XXXX

>>>>

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  word_document:
    reference_docx: template_paper.docx
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

In spite of more or less substantial mismatches between model results
and data, it is remarkable that most of the models capture the essential
features of the highly complex process of tree establishment, while not
having been parameterized with such data. We conclude that in spite of
all the uncertainties, a lot can be gained by a focus on the modeling of
establishment processes in these models, this will make their long-term
forest dynamics projections under changing environmental conditions much
more robust.

@@

More specifically, we aimed to answer the following questions: (1) Are
models of forest dynamics capturing accurately tree ingrowth levels,
initial tree diversity and mortality in tree establishment? (2) Do model
traits explain differences in model performance? (3) How well do models
capture total ingrowth and individual species regeneration niches across
environmental gradients of light availability, temperature, and soil
moisture?

@@

The starting point for the models establishment differs as well, ranging
from seedbank, seed or seedling to sapling (i.e., trees much larger than
10 cm height, often at least 2-3 m tall). Models that start from seed
need to include a larger number of ecological processes such as
germination and survival of young seedlings, whereas models that start
from saplings have to aggregate via parameterizations several ecological
processes because they are not treated explicitly. This reduces model
complexity but comes at the cost of blurring process representation. In
the models used in this study the overall complexity in the
establishment modules varies considerably. Following @bugmann2022, we
can classify seven models (iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS,
ForClim and TreeMig) as having rather high complexity in their
establishment modules (mean establishment formulation complexity
complexity across all processes \>0, Table \@ref(tab:tableModels)).

Two models, ForClim and ForCEEPS, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a different approach where
individual species properties and its relation to the environment are
more important [@huber2020]. These two model variants allow us to
evaluate a more process-based and complex module (variant 11) against a
simple module (variant 1), while the rest of the model structure is
identical. Similarly, the two ForCEEPS variants allow us to isolate the
importance of the canopy feedback (i.e., simulated actual composition
and relative abundance of species in the plot) via seed trees for the
quantity and quality of simulated ingrowth, as one variant includes this
feedback whereas the other does not.

@@

kable(modelDF, 
      format = "latex", booktabs = TRUE,
      caption = "Models included and their characteristics.") |>
        kableExtra::landscape() |>
        kableExtra::kable_styling(full_width = F, latex_options = "scale_down") |>
    # fix width by column
      kableExtra::column_spec(column = 1, width = "5em") |> 
      kableExtra::column_spec(column = 2, width = "3em") |> 
      kableExtra::column_spec(column = 3, width = "3em") |> 
      kableExtra::column_spec(column = 4, width = "4em") |>
      kableExtra::column_spec(column = 5, width = "5em") |>
      kableExtra::column_spec(column = 6, width = "5em") |>
      kableExtra::column_spec(column = 7, width = "5em") |>
      kableExtra::column_spec(column = 8, width = "8em") |>
      kableExtra::column_spec(column = 9, width = "8em") |>
      kableExtra::column_spec(column = 10, width = "5em") |>
      kableExtra::column_spec(column = 11, width = "5em") |>
      kableExtra::column_spec(column = 12, width = "8em") |>
      kableExtra::add_footnote(c("S-Stand, L-Landscape, G-Global",
                                 "PB-Process based, EM-Empirical", 
                                 "C-Cohort, I-Individual",
                                 "RG-Regeneration, RC-Recruitment",
                                 "Y-Yes, N-No",
                                 "D-Daily, M-Monthly"),

@@

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, water
balance, soil quality and exposition (i.e., slope and aspect). Each of
these sites featured at least two consecutive measurements. Ingrowth
thresholds for the sites differed between 0 and 10 cm, we defined two
data sets where one included all sites with a diameter threshold of 7 cm
or lower and the other included another 35 sites with diameter
thresholds between 7 and 10 cm. Thus we focused on two data sets with
the harmonized diameter thresholds of 7 cm and 10 cm.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species diversity of
the ingrowth, and ingrowth mortality. We evaluated the species diversity
across models and in relation to the observed data by calculating the
Shannon index based on the relative proportion of the species in terms
of basal area. The Shannon index was calculated for the establishment
($H_{R_{n}}$) (Equation \ref{eqpiShannonR}) and at the stand level
($H_{S_{n,s}}$)(Equation \ref{eqpiShannon}). The higher the value of the
index, the higher the species diversity at a particular site and sample.

```{=tex}
\begin{equation}
H_{R_{n}} = - \sum_{i = 1}^{s}{p_{R_{n,i}}\ln(p_{R_{n,i}})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{R_{n,i}} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{S_{n}} = - \sum_{i = 1}^{s}{p_{S_{n,i}}\ln(p_{S_{n,i}})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{S_{n,i}} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and have basal area
above zero in sample $n$; $p_{n,r,i}$ is the proportion of the species
$i$ in sample $n$ calculated as the ingrowth basal area ($r_{BA_{n,i}}$)
for that species $i$ relative to the recruited basal area
($TotalR_{BA_{n}}$) of the sample $n$; $p_{S_{n,i}}$ is the proportion
of the species $i$ calculated as the basal area of all trees
($BA_{n,i}$) of that species $i$ relative to the total basal area
($TotalR_{BA_{n}}$) of the sample $n$.

Mortality in tree establishment was assessed based on the ratio of
ingrowth between the 7 and 10 cm diameter thresholds. We used the
Reineke self-thinning rule [@reineke1933; @pretzsch2003] as a reference
to estimate whether the different ingrowth values were above or below
the expected theoretical rate between these two diameters. The Reineke
self-thinning rule is usually calculated from even-aged, single species
stands and assumes a relationship between the number of stems and the
quadratic mean diameter in fully stocked pure stands. The value used in
our comparisons was 1.77 (i.e., we expect stem numbers at 7 cm to be 77%
higher than at 10 cm), calculated using Equation \ref{eqReineke}.

```{=tex}
\begin{equation}
Reineke = (7/10)^{(-1.605)} \label{eqReineke}
\end{equation}
```
We assessed model traits in relation to model performance focusing on
features such as model complexity as defined in @bugmann2022, model type
(empirical or process based), models with and without canopy feedback,
and the models scale (stand, landscape or global). We tested for
significant differences using two sample t-tests by considering each of
these model features considering observed and simulated ingrowth
overestimation, ingrowth levels and species diversity respectively.

The total ingrowth and individual species regeneration niches were
evaluated across environmental gradients of light availability,
temperature, and soil moisture. Ingrowth levels were evaluated along
three key environmental gradients, i.e., stand basal area (the basal
area of all the trees in each sample including tree establishment) as a
proxy of light availability at the forest floor, the annual degree-day
sum [@allen1976; @fischlin1995] as a proxy for growing season warmth,
and the climatic water balance as a proxy for soil moisture
[@speich2019]. The mean ingrowth values were calculated as the mean
across the 200 samples per site. The observed data were modeled using a
Generalized Additive Model [@wood2011] with a negative binomial
distribution and restricted maximum likelihood to better understand the
environmental gradients trend in the levels of the observed ingrowth,
relative to the simulation results per model.

@@

```{r over, echo=FALSE, fig.cap="Mean ingrowth levels across all samples per site, plotted for the 200 sites and for each model. The red dashed lines shows the 25th and 75th percentiles for the 7 cm diameter threshold in the observed data. There are two boxplots for each model where the left lighted boxplot corresponds to 7 cm and the right darker boxplot to 10 cm", out.width='100%'}
        knitr::include_graphics(here::here("figures", "recOverAll710.png"))
```

@@

```{r H710, echo=FALSE, fig.cap="Shannon index of tree ingrowth, calculated by basal area, as the mean value across the 200 samples per site for the diameter thresholds of 7 and 10 cm, respectively. The red dashed lines shows the 25th and 75th percentiles for the 7 cm diameter threshold in the observed data.", out.width='100%'}
    knitr::include_graphics(here::here("figures", "H7_10.png"))
```

@@

For both the observed and the simulated data, and both ingrowth
thresholds, species diversity in the ingrowth compared to stand-level
diversity did not have clear patterns across all the models (Figures S1
and S2). Most models captured reasonably well or overpredicted species
diversity at the stand level (Figure \@ref(fig:combinedRichness)A, left
and center; cf. Figure S3), but a characteristic feature was that
simulations had a much lower variability of diversity compared to
observations. Models overpredicting species diversity included ForClim1,
ForClim 11, FORMIND and TreeMig (Figure \@ref(fig:combinedRichness)A,
left). Several models underpredicted species diversity, i.e., LPJ-GUESS,
ForCEEPS(f) and 4C (Figure \@ref(fig:combinedRichness)A, right).

The majority of the models overpredicted ingrowth diversity and only two
of them underestimated ingrowth diversity (4C and Landis-II) (Figure
\@ref(fig:combinedRichness)B and Figure S4). Both the models with
feedback and those without feedback from the adult trees to ingrowth
(via seed production) overestimate, underestimated and captured
reasonably well ingrowth species diversity compared to the observations.
Again, simulated diversity was considerably less variable than observed
diversity in the ingrowth.

@@

There were strong differences among the models regarding the mortality
rate between 7 and 10 cm (Figure \@ref(fig:mortality11)). Most of the
models featured a mortality rate significantly larger than the observed
data, thus at least partially compensating for the general
overestimation of ingrowth levels (Figure \@ref(fig:over) and Table S3);
six models did not feature a significant difference compared to the
observed data (ForCEEPS, FORMIND, ForClim 1 and 11, iLand and aDGVM2).
Four models (xComp, PICUS, LandClim and TreeMig) featured very high
mortality rates (i.e., well above the Reineke self-thinning line), which
compensated for the strong initial overestimation of ingrowth (cf.
Figure \@ref(fig:over)). Interestingly, two models that underestimated
overall ingrowth levels (SIBYLA and Landis-II; cf. Figure
\@ref(fig:over)) featured mortality rates that were close to but still
above the self-thinning line (Figure \@ref(fig:mortality11)).

@@

There was no significant relationship between the mean complexity of the
establishment module (Table \@ref(tab:tableModels)) and the
overestimation proportion at a diameter threshold of 7 cm (Figure S6).
All the models except 4C had significant differences between the
observed mean ingrowth values and the simulated, therefore it was not
possible to differentiate in terms of performance between models forest
type (empirical or process based) and across models scales (stand,
landscape or global) (Table S5). The ingrowth species diversity values
simulated per models were also assessed againts the observed data to see
if there was a difference between models that include feedback compare
to those that did not include feedback, all the models except
ForCEEPs(f) and FORMIND had significant differences (Table S6). It is
noteworthy the change from overpredicted species diversity in the
ingrowth for the model ForCEEPS to a closer diversity levels when the
recruitment module included feedback (ForCEEPs(f)).

@@

The observed ingrowth did not change much across the climatic water
balance gradient, and showed a slight increase with temperature. The
models did not match the observed ingrowth patterns across gradients of
soil moisture and temperature (Figure \@ref(fig:nvTrend710BAlightTemp)
and S7), and neither was there a clear pattern across models. Regarding
the soil moisture gradient, a group of models featured decreasing
ingrowth with increasing soil moisture (4C, ForClim 1, ForClim11,
SIBYLA, xComp, PICUS, LandClim, TreeMig and iLand), while a few models
showed the opposite trend (ForCEEPS(f), FORMIND and LPJ-GUESS). Across
the temperature gradient, all the models that featured a decrease in
ingrowth levels with increasing soil moisture showed the opposite trend
(i.e., more ingrowth with increasing temperature), except for 4C. The
other models did not feature a clear pattern.

@@

Correctly capturing species composition in tree establishment is
important to assess the future functional diversity of a forest, e.g.,
its sensitivity to drought or resilience to disturbances [@redmond2015;
@seidl2022]. In the simulations, overall ingrowth species diversity
levels were well within the observed range for half of the models (7 out
of 14). Thus, while most of the models are facing difficulties to
quantitatively match ingrowth levels (cf. above), their performance is
better when it comes to matching the composition of ingrowth as a
function of abiotic and biotic conditions. Most models maintained or
even decreased ingrowth diversity between the 7 and 10 cm thresholds,
and the same is visible from the empirical data, but the differences are
not significant. Diversity variations may be explained by the way the
individual models consider establishment processes [@konig_tree_2022],
e.g., in the absence of adults, ingrowth can be linked to either seed
rain-dispersal, or to a seed/seedling bank.

The species diversity of the entire stand was better captured by the
models than the diversity of tree ingrowth, and only a few models
overpredicted stand-level diversity. Defining ingrowth composition as
being proportional to adult abundance, regardless of productivity, might
be a simple and conservative assumption to safely maintain relative
species abundances [@hanbury-brown2022], although this may be
simplistic. Furthermore, based on the simulation results, there is no
evidence that models with feedback captured better the ingrowth species
diversity than those without feedback with the exception of ForCEEPS
that significantly improved the ingrowth species diversity with the
model version that included feedback (ForCEEPS(f)). The similar
performance of models with and without feedback is likely because the
models put more weight to the regeneration niche arising from abiotic
and biotic filters, than from the habitat niche of the adult trees
[@grubb1977].

@@

Funding of the workshop was provided by the COST Action CA19139 PROCLIAS
(PROcess-based models for CLimate Impact Attribution across Sectors),
supported by COST (European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Grant ETH-35 18-1). ODY was funded by SNF project XXXX. KB received
support from the European Research Council under the European Union's
Horizon 2020 research and innovation program (Grant Agreement
101001905). JB acknowledges funding from the long-term research
development project no. RVO 67985939 from the Czech Academy of Sciences.
MM acknowledges financial support from I-Maestro (Innovative forest
management strategies for a resilient bioeconomy under climate change
and disturbances, grant no. 22035418, 2019--2022) funded by the ERA-NET
Cofund ForestValue. KM received funding from the Project No.
CZ.02.1.01/0.0/0.0/16_019/0000803 financed by OP RDE, grant no.
ITMS2014+313011W580 supported by the Integrated Infrastructure
Operational Programme funded by the ERDF
